The promise of community-driven preprints in ecology and evolution

### 2024-06-20 ms/ms.qmd
Update format for Proc B and adding in a abstract

---
title: "The promise of community-driven preprints in ecology and evolution"
bibliography: ../bib/refs.bib
csl: ../bib/nature.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

# Introduction

@@

# Getting to know your *EcoEvoRxiv* preprint server

@@

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published on *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the launch in November 2018. Notable milestones include *EcoEvoRxiv* transitioning to the California Digital Library (CDL), the acceptance of preprints and postprints in Spanish and Portuguese, and the acceptance of the first IUCN Red List Ecosystem report; B) Geographic origin of the preprints and postprints uploaded to *EcoEvoRxiv*, inferred from the country of affiliation of the submitting author; C) Taxa used/covered in the articles posted to *EcoEvoRxiv* (n = 1080 articles); D) Types of preprints accepted on *EcoEvoRxiv* (n = 620 articles). E) Academic age of authors posting to *EcoEvoRxiv* along with the gender of the submitting author. Values lower than zero are indicative of authors who uploaded preprints before their first scientific publication in a journal. Map base source: R Package "maps" v.3.4.2. Shapefile: Natural Earth https://www.naturalearthdata.com/about/terms-of-use/.

@@

### *Overview of **EcoEvoRxiv** preprints (and postprints)*

@@

### *Diversifying article types on **EcoEvoRxiv**: overcoming the 'grey literature' problem*

@@

### *Breaking down language barriers to scientific communication: improving diversity and data representation globally*

@@

### *Generational and gender-based gaps in preprinting practices*

@@

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Summary of the publication status of articles on *EcoEvoRxiv*. A) Time between uploading an preprint to *EcoEvoRxiv* and its publication as a peer-reviewed journal article. Articles that were published within 2 months (60 days) of being submitted are considered postprints and are not included in this figure. B) Access status of articles linked to *EcoEvoRxiv* preprints classified as “Open Access” or “Not Open Access”. “Unknown” status is for articles not explicitly identified as being open access or not on the Unpaywall platform. C) Sub-types of open access status of articles linked to *EcoEvoRxiv* preprints. For full details on the meaning of each category see  https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean-.

@@

### *Science takes time, but publication could take longer*
Increased competition in science has raised the bar with respect to the amount of data required for publication [@vale2015accelerating]. This requirement is a good outcome if it results in higher-impact research that better clarifies our understanding of the natural world, but it does come at a cost for the speed of research dissemination [@vale2015accelerating]. Preprints have been proposed as a way to disseminate research more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is needed to quantify the real benefit of preprints in this context.

@@

### *Cautious 'open'-mindedness of research in preprints*
In addition to speeding up dissemination, preprints and postprints can also be a useful way to ensure that research remains open and accessible to the research community irrespective of the accessibility of the final peer-reviewed paper [@bourne2017ten; @vale2015accelerating]. We evaluated whether articles hosted at *EcoEvoRxiv*, and that were also published in a journal, were published open access. The open access status of each published article was obtained using the R package *roadoi* (v.`r utils::packageVersion("roadoi")`) to connect to the Unpaywall platform [@jahn2024]. Most of the published articles were open access [`r sum_is_oa$valid_percent[2]*100`% (*n* = `r sum_is_oa$n[2]` out of `r sum(sum_is_oa$n[1:2])` where the status was known); [@fig-pubsummary]A]; however, `r sum_is_oa$valid_percent[1]*100`% (*n* = `r sum_is_oa$n[1]`) were published behind paywalls. For articles published in open access journals, the type of open access also varied widely (e.g., Gold, Hybrid, Green OA etc., [@fig-pubsummary]B).

@@

# Paving our future to open, transparent and community-driven science

@@

# Acknowledgements
We would like to thank the California Digital Library (CDL) and the CDL team (particularly, Alainna Wrigley, Justin Gonder, Lisa Schiff, Catherine Mitchell, Hardy Pottinger and Amanda Karby) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE). We would like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Committee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2023 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper. DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276). SN and ML are supported by the Australian Research Council (ARC) Discovery Project Grants (DP210100812 and DP230101248).

# Data and Code Availability
All data and code can be found on GitHub at: https://github.com/daniel1noble/ecoevo_1000

# Conflict of Interest
The authors would like to acknowledge competing interests on the perspectives presented in this paper given that many (DWAN, SN, ML) are founding members of *EcoEvoRxiv* and/or are part of the *EcoEvoRxiv* committee.

>>>>

---
title: "The promise of community-driven preprints in ecology and evolution"
bibliography: ../bib/refs.bib
csl: ../bib/proceedings-of-the-royal-society-b.csl
format:
  docx:
    reference-doc: ../bib/template.docx
  pdf:
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "figure"
  tbl-prefix: "table"
---

@@

# Abstract
Publishing preprints has become an entrenched practice across a multitude of scientific disciplines and it is quickly becoming commonplace in ecology and evolutionary biology. Preprints can facilitate the rapid sharing of scientific knowledge to establish precedence, speed up the dissemination of research findings, and enable feedback from the research community before peer review. Yet, significant barriers to preprint use exist including language barriers, a lack of understanding about the benefits of preprints and a lack of diversity in the types of research outputs accepted (e.g., reports). Community driven preprint initiatives can allow a research community to come together to break down these barriers and move scientific publishing practices into new and exciting directions that promote greater equity and better coverage of global knowledge.  Here, we explore the first preprints uploaded to *EcoEvoRxiv*, a community-driven preprint server for ecologists and evolutionary biologists, to characterise preprint practices in ecology, evolution and conservation. Our perspective piece highlights some of the unique initiatives that *EcoEvoRxiv* has taken to break down barriers to scientific publishing by exploring the composition of articles, how gender and career stage influence preprint use, whether preprints are associated with greater open science principles (e.g., code and data sharing), and tracking preprint publication outcomes. Our analysis identifies areas that we still need to improve upon but highlight how community-driven initiatives, such as *EcoEvoRxiv*, can play a crucial role in shaping publishing practices in biology.

# 1. Introduction

@@

# 2. Getting to know your *EcoEvoRxiv* preprint server

@@

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: "Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published on *EcoEvoRxiv* between 2018 and 2023. *EcoEvoRxiv* was established in June 2018 before the launch in November 2018. Notable milestones include *EcoEvoRxiv* transitioning to the California Digital Library (CDL), the acceptance of preprints and postprints in Spanish and Portuguese, and the acceptance of the first IUCN Red List Ecosystem report; B) Geographic origin of the preprints and postprints uploaded to *EcoEvoRxiv*, inferred from the country of affiliation of the submitting author; C) Taxa used/covered in the articles posted to *EcoEvoRxiv* (n = 1080 articles); D) Types of preprints accepted on *EcoEvoRxiv* (n = 620 articles). E) Academic age of authors posting to *EcoEvoRxiv* along with the gender of the submitting author. Values lower than zero are indicative of authors who uploaded preprints before their first scientific publication in a journal."

@@

### (a) *Overview of **EcoEvoRxiv** preprints (and postprints)*

@@

### (b) *Diversifying article types on **EcoEvoRxiv**: overcoming the 'grey literature' problem*

@@

### (c) *Breaking down language barriers to scientific communication: improving diversity and data representation globally*

@@

### (d) *Generational and gender-based gaps in preprinting practices*

@@

# 3. Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Summary of the publication status of articles on *EcoEvoRxiv*. A) Time between uploading an preprint to *EcoEvoRxiv* and its publication as a peer-reviewed journal article. Articles that were published within 2 months (60 days) of being submitted are considered postprints and are not included in this figure. B) Access status of articles linked to *EcoEvoRxiv* preprints classified as “Open Access” or “Not Open Access”. “Unknown” status is for articles not explicitly identified as being open access or not on the Unpaywall platform. C) Sub-types of open access status of articles linked to *EcoEvoRxiv* preprints. For full details on the meaning of each category see  https://support.unpaywall.org/support/solutions/articles/44001777288-what-do-the-types-of-oa-status-green-gold-hybrid-and-bronze-mean-.

@@

### (a) *Science takes time, but publication could take longer*
Increased competition in science has raised the bar with respect to the amount of data required for publication [@vale2015accelerating]. This requirement is a good outcome if it results in higher-impact research that better clarifies our understanding of the natural world, but it does come at a cost for the speed of research dissemination [@vale2015accelerating]. Preprints have been proposed as a way to disseminate research more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is needed to quantify the real benefit of preprints in this context.

@@

### (b) *Cautious 'open'-mindedness of research in preprints*
In addition to speeding up dissemination, preprints and postprints can also be a useful way to ensure that research remains open and accessible to the research community irrespective of the accessibility of the final peer-reviewed paper [@bourne2017ten; @vale2015accelerating]. We evaluated whether articles hosted at *EcoEvoRxiv*, and that were also published in a journal, were published open access. The open access status of each published article was obtained using the R package *roadoi* (v.`r utils::packageVersion("roadoi")`) to connect to the Unpaywall platform [@jahn2024]. Most of the published articles were open access [`r sum_is_oa$valid_percent[2]*100`% (*n* = `r sum_is_oa$n[2]` out of `r sum(sum_is_oa$n[1:2])` where the status was known); [@fig-pubsummary]A]; however, `r sum_is_oa$valid_percent[1]*100`% (*n* = `r sum_is_oa$n[1]`) were published behind paywalls. For articles published in open access journals, the type of open access also varied widely (e.g., Gold, Hybrid, Green OA etc., [@fig-pubsummary]B).

@@

# 4. Paving our future to open, transparent and community-driven science

@@

# Data accessibility
All data and code can be found on GitHub at: https://github.com/daniel1noble/ecoevo_1000

# Competing interests
The authors would like to acknowledge competing interests on the perspectives presented in this paper given that many (DWAN, SN, ML) are founding members of *EcoEvoRxiv* and/or are part of the *EcoEvoRxiv* committee.

# Funding
DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276). SN and ML are supported by the Australian Research Council (ARC) Discovery Project Grants (DP210100812 and DP230101248).

# Acknowledgements
We would like to thank the California Digital Library (CDL) and the CDL team (particularly, Alainna Wrigley, Justin Gonder, Lisa Schiff, Catherine Mitchell, Hardy Pottinger and Amanda Karby) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolutionary Biology (SORTEE). We would like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Committee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2023 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper.
